Supp mat: time series paper

### 2017-05-15 writing/manuscript_draft.Rmd
added related work in discussion

- intro long time series
- my general question
- primer on power
- more general power interpretation
- 2+ approaches
- what we do in this paper

Observational studies, and censusing populations, has become a cornerstone of modern ecology and conservation biology [@Magurran2010]. Longterm data is necessary to understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time. There are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

@@

Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..

@@

We show two different approaches to estimate the minimum time series length required required to address a particular question of interest. The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987]. We focus on a empirical approach to estimate the minimum time series length requried.

The first question we asked was, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary. In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation.

@@

@@

- greater than 5 years [@Morrison2008]
- more than 10-15 years [@Rueda-Cediel2015]
- general advice (variability, correlations) [@Rhodes2011]
- more than 10 years [@Gerber1999]
- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]

>>>>

<!--
- intro long time series
- my general question
- primer on power
- more general power interpretation
- 2+ approaches
- what we do in this paper
-->

Observational studies and population censuses have become a cornerstone of modern ecology and conservation biology [@Magurran2010]. Longterm data is necessary to understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time. There are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

@@

Past work has investigated these questions with either theoretical approaches not using data or by examining a single population. As an example of the former, [@Rhodes2011] examined..
- need to expand this paragraph in line with what I wrote in the discussion...

@@

We show two different approaches to estimate the minimum time series length required required to address a particular question of interest. The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987]. Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.

First we asked, what is $t_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression? We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig. \ref{fig:fig:min_time_dist}). In line with theoretical predictions (citations), we also found $t_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).

@@

@Rueda-Cediel2015 found that 10-15 years was adequate to characterize population trends in a short lived animal. They used a matrix model of a particular species...

@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species. They used a longterm census to retroactively determine the minimum time series required to assess threat status. They found that only XX years were needed, XX years before the delisting decision was actually made. This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].

There has also been interest in determining the minimum time series length in the bird community (citations). (Bart 2004, Hatch 2003) [@Seavy2007] Hatch 2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design (cite Hatch). Similarly, Bart et al 2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing longterm population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautionsioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..

@@

- greater than 5 years [@Morrison2008]
- more than 10-15 years [@Rueda-Cediel2015]
- general advice (variability, correlations) [@Rhodes2011]
- more than 10 years [@Gerber1999]
- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
- paper on syncrony from James

## Supporting Information

- list each bit of supporting information. Github reference

## Acknowledgements

ERW was partially supported by a National Science Foundation Graduate Fellowship. We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight. We would also like to thank X anonymous reviewers for their helpful comments.

### 2017-05-15 writing/manuscript_draft.tex
added related work in discussion

Observational studies, and censusing populations, has become a
cornerstone of modern ecology and conservation biology (Magurran et al.
2010). Longterm data is necessary to understand population dynamics and
to assess extinction risk of species. Even though many time series may
now be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. There are
of course notable exceptions including the Isle Royale study, Iowa corn
study, cod, lynx-hair, and many others.

@@

Past work has investigated these questions with either theoretical
approaches not using data or by examining a single population. As an
example of the former, (Rhodes and Jonzen 2011) examined..

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). We focus on a empirical approach to estimate
the minimum time series length requried.

The first question we asked was, what is \(t_{min}\) for determining
whether a population was increasing or decreasing over time according to
linear regression. We found that for 868 populations, at least 10-15
years of continuous monitoring were neccessary. In line with theoretical
predictions (citations), we also found \(t_{min}\) was strongly
correlated with the trend strength, variability in population size, and
temporal autocorrelation.

@@

@@

\begin{itemize}
\tightlist
\item
  greater than 5 years (Morrison and Hik 2008)
\item
  more than 10-15 years (Rueda-Cediel et al. 2015)
\item
  general advice (variability, correlations) (Rhodes and Jonzen 2011)
\item
  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
\item
  specific recommendations depend on amount of power and strength of
  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
\end{itemize}

>>>>

Observational studies and population censuses have become a cornerstone
of modern ecology and conservation biology (Magurran et al. 2010).
Longterm data is necessary to understand population dynamics and to
assess extinction risk of species. Even though many time series may now
be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. There are
of course notable exceptions including the Isle Royale study, Iowa corn
study, cod, lynx-hair, and many others.

@@

Past work has investigated these questions with either theoretical
approaches not using data or by examining a single population. As an
example of the former, (Rhodes and Jonzen 2011) examined.. - need to
expand this paragraph in line with what I wrote in the
discussion\ldots{}

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). Here, we focus on a empirical approach to
estimate the minimum time series length requried to address a given
question.

First we asked, what is \(t_{min}\) for determining whether a population
was increasing or decreasing over time according to linear regression?
We found that for 868 populations, at least 10-15 years of continuous
monitoring were neccessary (Fig. \ref{fig:fig:min_time_dist}). In line
with theoretical predictions (citations), we also found \(t_{min}\) was
strongly correlated with the trend strength, variability in population
size, and temporal autocorrelation (cite figure).

@@

Rueda-Cediel et al. (2015) found that 10-15 years was adequate to
characterize population trends in a short lived animal. They used a
matrix model of a particular species\ldots{}

Gerber, DeMaster, and Kareiva (1999) investigated the minimum time
series required to estimate population growth of an endagered, but
recovering, species. They used a longterm census to retroactively
determine the minimum time series required to assess threat status. They
found that only XX years were needed, XX years before the delisting
decision was actually made. This highlights the importance of estimating
the minimunm time series required as an earlier decision would have
saved time and money (Gerber, DeMaster, and Kareiva 1999).

There has also been interest in determining the minimum time series
length in the bird community (citations). (Bart 2004, Hatch 2003) (Seavy
and Reynolds 2007) Hatch 2003 used seabird monitoring data to estimate
the minimum time needed to achieve appropriate levels of power. He found
that the minimum time required ranged between 11 and 69 years depending
on species, trend strength, and study design (cite Hatch). Similarly,
Bart et al 2004 recommended a standard of 80\% power to detect a 50\%
decline over 20 years in landbird populations. They found that only
about 42\% of populations in the Breeding Bird Survey met their
criteria. Seavy and Reynolds (2007) asked whether statistical power was
even a useful framework for assessing longterm population trends. They
used 24 years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulting in high statistical power
(above 0.8). Therefore, they cautionsioned against only using power
analyses to design monitoring schemes and instead argued for precision
based metrics ..

@@

\begin{itemize}
\tightlist
\item
  greater than 5 years (Morrison and Hik 2008)
\item
  more than 10-15 years (Rueda-Cediel et al. 2015)
\item
  general advice (variability, correlations) (Rhodes and Jonzen 2011)
\item
  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
\item
  specific recommendations depend on amount of power and strength of
  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
\item
  paper on syncrony from James
\end{itemize}

\subsection{Supporting Information}\label{supporting-information}

\begin{itemize}
\tightlist
\item
  list each bit of supporting information. Github reference
\end{itemize}

\subsection{Acknowledgements}\label{acknowledgements}

ERW was partially supported by a National Science Foundation Graduate
Fellowship. We would like to thank members of the Ecological Theory
group at the University of California, Davis for their insight. We would
also like to thank X anonymous reviewers for their helpful comments.
